Ensemble Methods for Robust Feature Selection

### 2016-08-05 paper/paper.tex

\section{Feature selection methods}

Symmetrical Uncertainty (SU) \citep{press1996numerical} is an entropy based measurement. It is the normalized mutual information between the feature and the class. The mutual information measures the influence of the knowledge of the class on the feature value in terms of entropy and normalizes it. It is defined as follows:
\begin{equation}
  \label{eq:su}
  SU(F,C) = 2 \frac{H(F) - H(F|C)}{H(F) + H(C)} \textrm{, where } H(\cdot) \textrm{ is the entropy.}
\end{equation}
More details about mutual information can be found in \cite{paninski2003estimation}.

The second feature selection is RELIEFF \citep{kononenko1997overcoming}. For each sample $x_i$ the k nearest neighboring samples, using the $L_1$ norm as metric, belonging to the same class $\{\textrm{Near-hit}_{i, j}\}_{j=1..k}$ and the ones of the opposite class $\{\textrm{Near-miss}_{i, j}\}_{j=1..k}$ are selected. Then the difference is taken to update the weight vector representing how well each feature distinguishes samples of the two classes, as can be seen in equation \ref{eq:relief}. In all the experiments k is set to 5, which empirically gives satisfactory results.
\begin{equation}
  \label{eq:relief}
  W_i = W_i - \frac{1}{k}\sum_{j=1}^{k}|x_i - \textrm{Near-hit}_{i,j}| + \frac{1}{k}\sum_{j=1}^{k} |x_i - \textrm{Near-miss}_{i,j}|
\end{equation}
The third feature selection method uses recursive feature elimination (RFE) with a linear C-SVM classifier \citep{guyon2002gene} to weight the features. 
The RFE method fits a classifier and ranks the feature according to their importance in the model. In the case of the C-SVM, it is based on the weight vector of the hyperplane. More precisely after obtaining the solution for the primal problem of the C-SVM in equation \ref{eq:svm_primal}, the ranking coefficients $c_i$ for the features are then calculated by squaring the weights $c_i = w_i^2$.
\begin{alignat}{-1}
     \min_{w,b}  & \quad \frac12\|w\|^2 + C\sum_{i=1}^m \xi_i & & & \label{eq:svm_primal} \\ 
   \text{s.\,t.} & \quad y^{(i)} (w^Tx^{(i)} + b) & \quad \geq & \quad 1 -\xi_i &&
                   \quad \forall \, i = 1, \ldots, m \nonumber \\
                 & \quad \xi_i                  & \quad \geq & \quad 0 &&
                   \quad \forall \, i = 1, \ldots, m \nonumber 
\end{alignat}
The worst ranked features are then discarded and the procedure is repeated until only a subset of features is left. In the experiments, $10\%$ of the size of all features are discarded at each step. The procedure is repeated until a subset of the size of $1\%$ of all features is left. Since in each step $10\%$ of all features are discarded, the overall process takes $10$ steps. Therefore a C-SVM is needed to be $10$ times evaluated. Another method would be to discard only $10\%$ of the remaining features. A more gradually ascend of feature weights can be achieved this way. However, it is not used due to the substantially increase in needed computation power. For the hpyerparameter $C$ a grid search on the set $\{1, 10, 100, 1000\}$ is done. The parameters are evaluated with stratified 5-fold cross-validation. In this work the RFE method with a C-SVM classifier is abbreviated as SVM RFE.

@@

\subsection{Data Sets}
Datasets are taken from the NIPS 2003 challenge (\cite{NIPS}). However, due to computation power constraints, not the whole dataset is always used. Additionally the colon dataset (\cite{alon1999broad}) is used. Finally, artificial data is also generated, since here it is possible to control which features are meaningful. All data sets are binary labelled.

\begin{center}
    \begin{tabular}{| r | c | c | c | c | c |}
    \hline
    Name & \# Features & \# Samples & \# Class 1 & \# Class 2 & Ref\\ \hline
    Arcene & 10 000 & 200 & 112 & 88 & \cite{NIPS} \\
    Dexter & 20 000 & 600 & 300 & 300 & \cite{NIPS} \\
    Gisette & 5000 & 200 & 92 & 108 & \cite{NIPS} \\
    Colon & 2000 & 62 & 40 & 22 &  \cite{alon1999broad} \\
    Artificial & 10 000 & 300 & 150 & 150 & \\
    \hline
    \end{tabular}
\end{center}

The features of the artificial dataset is generated with different Gaussian distributions, depending on whether they are used in the labeling process or not. It consists of $100$ significant features and $9900$ fake features. The significant features and fake features both consists of correlated and uncorrelated features. The correlation between the correlated features is randomly generated. The ratio of correlated features and uncorrelated features are for the significant and the fake features $0.5$. 
\begin{align}
f \sim \mathcal{N}(0, 1) + \mathcal{N}(0, \Sigma)  \quad \text{with} \quad \Sigma_{ij} \sim \mathcal{U}(0,1)
\end{align}
Therefore the artificial consists of $50$ correlated significant features, $50$ uncorrelated significant features, $4950$ correlated fake features and $4950$ uncorrelated fake features. In this work, the uncorrelated features and the correlated features are also respectively called univariate and  multivariate features.
To generate balanced binary labels the significant features are first summed, then the median of the sum is subtracted, finally the sign is taken.

@@

The model performance is estimated by measuring the balanced error rate of a set of classifiers: Lasso logistic regression, linear C-SVM, k-nearest neighbor, Random Forest. 
%TODO describe how hyperparameters are obtained for each classifier
For each classifier the performance is assessed using a 10-fold cross validation setting and trained on the best 1\% of the features.

To measure the robustness, 10 equally sized subsamples are drawn from the dataset. The feature selector runs on all of them, and the best ones are taken. The subsets are then compared with a similarity measure. Given that the model performance is measured on the best 1\% of the features, the order of the feature should not be of importance but only if one is or is not part of the best subset. The Jaccard Index is used because only how many features are common to both sets is important. For two subsets of features $F_i$ and $F_j$ it is defined as:

@@

Furthermore, the feature selectors are evaluated with respect to how many significant features they capture in the $100$ the highest weighted features. RELIEFF achieves the highest precision of $0.5$, SVM RFE achieves $0.48$, then Lasso with $0.24$ and SU with $0$.

@@

To compare the different feature selection methods in relation of robustness and accuracy, the single feature selection and the three ensemble methods achieving the best results are plotted with accuracy against stability. For the stability the Jaccard index with $1\%$ of the features is used. For the accuracy the all classifiers described in \ref{sec:experiments}  reused. For each feature selection method the standard deviation for the accuracy and the stability on the different subsets are plotted. The results for the different classifiers are similar. In figure \ref{fig:boxplot_svm} the results for the C-SVM classifier can be observed. The other results can be seen in the appendix. %TODO boxplots for plots
As can be seen in figure \ref{fig:boxplot_svm} the ensemble methods achieve good results on each dataset. While single feature selection methods achieve better results in terms of accuracy and robustness on some datasets, they fail on others. However the ensemble methods keep their performance for each dataset. On the arcene dataset they even achieve better results than any single feature selector.

\begin{figure}[H]
  \centering
    \includegraphics[width=\textwidth]{images/boxplot_svm.pdf}
  \caption{The different ensemble methods compared with single feature selectors in comparison of stability and accuracy.
  The gray area is the accuracy when using no feature selection.}
  \label{fig:boxplot_svm}
\end{figure}

@@

\appendix
\label{appendix}
\section{Python libraries}
\label{apd:libraries}
For the feature selection methods SU and RELIEFF come from the scikit-feature library (\cite{Li-etal16}).
For C-SVM, RFE, Lasso logistic regression come from the scikit-sklearn library (\cite{scikit-learn}).
\section{Additional plots}
\label{apd:plots}
\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{images/feature_weights_hist.png}
  \caption{The histogram of the feature weights for all feature selectors used for all datasets used.}
  \label{fig:feature_weights_hist}
\end{figure}

>>>>

\section{Feature selection methods}
\label{sec:fs_methods}
Symmetrical Uncertainty (SU) \citep{press1996numerical} is an entropy based measurement. It is the normalized mutual information between the feature and the class. The mutual information measures the influence of the knowledge of the class on the feature value in terms of entropy and normalizes it. It is defined as follows:
\begin{equation}
  \label{eq:su}
  SU(F,C) = 2 \frac{H(F) - H(F|C)}{H(F) + H(C)} \textrm{, where } H(\cdot) \textrm{ is the entropy.}
\end{equation}
More details about mutual information can be found in \cite{paninski2003estimation}.

The second feature selection is RELIEFF \citep{kononenko1997overcoming}. For each sample $x_i$ the k nearest neighboring samples, using the $L_1$ norm as metric, belonging to the same class $\{\textrm{Near-hit}_{i, j}\}_{j=1..k}$ and the ones of the opposite class $\{\textrm{Near-miss}_{i, j}\}_{j=1..k}$ are selected. Then the difference is taken to update the weight vector representing how well each feature distinguishes samples of the two classes, as can be seen in equation \ref{eq:relief}. In all the experiments k is set to 5, which empirically gives satisfactory results.
\begin{equation}
  \label{eq:relief}
  W_i = W_i - \frac{1}{k}\sum_{j=1}^{k}|x_i - \textrm{Near-hit}_{i,j}| + \frac{1}{k}\sum_{j=1}^{k} |x_i - \textrm{Near-miss}_{i,j}|
\end{equation}
The third feature selection method uses recursive feature elimination (RFE) with a linear C-SVM classifier \citep{guyon2002gene} to weight the features. 
The RFE method fits a classifier and ranks the feature according to their importance in the model. In the case of the C-SVM, it is based on the weight vector of the hyperplane. More precisely after obtaining the solution for the primal problem of the C-SVM in equation \ref{eq:svm_primal}, the ranking coefficients $c_i$ for the features are then calculated by squaring the weights $c_i = w_i^2$.
\begin{alignat}{-1}
     \min_{w,b}  & \quad \frac12\|w\|^2 + C\sum_{i=1}^m \xi_i & & & \label{eq:svm_primal} \\ 
   \text{s.\,t.} & \quad y^{(i)} (w^Tx^{(i)} + b) & \quad \geq & \quad 1 -\xi_i &&
                   \quad \forall \, i = 1, \ldots, m \nonumber \\
                 & \quad \xi_i                  & \quad \geq & \quad 0 &&
                   \quad \forall \, i = 1, \ldots, m \nonumber 
\end{alignat}
The worst ranked features are then discarded and the procedure is repeated until only a subset of features is left. In the experiments, $10\%$ of the size of all features are discarded at each step. The procedure is repeated until a subset of the size of $1\%$ of all features is left. Since in each step $10\%$ of all features are discarded, the overall process takes $10$ steps. Therefore a C-SVM is needed to be $10$ times evaluated. Another method would be to discard only $10\%$ of the remaining features. A more gradually ascend of feature weights can be achieved this way. However, it is not used due to the substantially increase in needed computation power. For the hpyerparameter $C$ a grid search on the set $\{1, 10, 100, 1000\}$ is done. The parameters are evaluated with stratified 5-fold cross-validation. In this work the RFE method with a C-SVM classifier is abbreviated as SVM RFE and sometimes SVM when it should be clear from the context.

@@

\subsection{Data Sets}
Datasets are taken from the NIPS 2003 challenge (\cite{NIPS}). However, due to computation power constraints, not the whole dataset is always used. Additionally the colon dataset (\cite{alon1999broad}) is used. Finally, artificial data is also generated, since here it is possible to control which features are meaningful. The datasets and the number of samples used are indicated in table \ref{table:datasets}. All data sets are binary labelled.

\begin{center}
    \begin{tabular}{| r | c | c | c | c | c |}
    \hline
    Name & \# Features & \# Samples & \# Class 1 & \# Class 2 & Ref\\ \hline
    Arcene & 10 000 & 200 & 112 & 88 & \cite{NIPS} \\
    Dexter & 20 000 & 600 & 300 & 300 & \cite{NIPS} \\
    Gisette & 5000 & 200 & 92 & 108 & \cite{NIPS} \\
    Colon & 2000 & 62 & 40 & 22 &  \cite{alon1999broad} \\
    Artificial & 10 000 & 300 & 150 & 150 & \\
    \hline
    \end{tabular}
  \label{table:datasets}
  \captionof{table}{Datasets used in the experiments}
\end{center}

The features of the artificial dataset is generated with different Gaussian distributions, depending on whether they are used in the labeling process or not. It consists of $100$ significant features and $9900$ fake features. The significant features and fake features both consists of correlated and uncorrelated features. The correlation between the correlated features is randomly generated. The ratio of correlated features and uncorrelated features are for the significant and the fake features $0.5$. 
\begin{alignat}{-1}
\text{uncorrelated features} & \sim & \mathcal{N}(0, 1) & \\ 
\text{correlated features} & \sim & \mathcal{N}(0, \Sigma) & \quad \text{with} \quad \Sigma_{ij} \sim \mathcal{U}(0,1)
\end{alignat}
Therefore the artificial consists of $50$ correlated significant features, $50$ uncorrelated significant features, $4950$ correlated fake features and $4950$ uncorrelated fake features. In this work, the uncorrelated features and the correlated features are also respectively called univariate and  multivariate features.
To generate balanced binary labels the significant features are first summed, then the median of the sum is subtracted, finally the sign is taken.

@@

The model performance is estimated by measuring the balanced error rate of a set of classifiers: Lasso logistic regression, linear C-SVM, k-nearest neighbor, Random Forest. The hyperparameter for Lasso logistic regression and linear C-SVM are obtained in the same way as for the feature selection in section \ref{sec:fs_methods}. For k-nearest neighbor 3 neighbor were used. For Random Forest we used 1 tree with a maximal depth of 5.
For each classifier the performance is assessed using a 10-fold cross validation setting and trained on the best $1\%$ of the features.

To measure the robustness, 10 equally sized subsamples are drawn from the dataset. The feature selector runs on all of them, and the best ones are taken. The subsets are then pairwise over all combinations compared with a similarity measure. Given that the model performance is measured on the best 1\% of the features, the order of the features should not be of importance but only if one feature is or is not part of the best subset. To measure the similarity between two subsets of features, the amount of features, which are in both subsets, is only important. Therefore we use the Jaccard Index as robustness measurement. For two subsets of features $F_i$ and $F_j$ it is defined as:

@@

Furthermore, the feature selectors are evaluated with respect to how many significant features they capture in the $100$ highest weighted features. RELIEFF achieves the highest precision of $0.5$, SVM RFE achieves $0.48$, then Lasso with $0.24$ and SU with $0$.

@@

To compare the different feature selection methods in relation of robustness and accuracy, the single feature selection and the three ensemble methods achieving the best results are plotted with accuracy against stability. For the stability the Jaccard index with $1\%$ of the features is used. For the accuracy the all classifiers described in \ref{sec:experiments} are reused. For each feature selection method the standard deviation for the accuracy and the stability on the different subsets are plotted. The results for the different classifiers are similar. In figure \ref{fig:boxplot_svm} the results for the C-SVM classifier can be observed. The other results can be seen in the appendix \ref{apd:plots} in figure \ref{fig:boxplots}.
As can be seen in figure \ref{fig:boxplot_svm} the ensemble methods achieve good results on each dataset. While single feature selection methods achieve better results in terms of accuracy and robustness on some datasets, they fail on others. However the ensemble methods keep their performance for each dataset. On the arcene dataset they even achieve better results than any single feature selector. On the artificial dataset all classifiers achieve bette results with feature selection than without. For the other datasets the C-SVM classifier without feature selection achieve better results except of colon where similar results are achieved. Furthermore the random feature selection method does also achieve good results in terms of accuracy on the datasets colon, arcene and gisette.

\begin{figure}[H]
  \centering
    \includegraphics[width=\textwidth]{images/boxplot_svm.pdf}
  \caption{The different ensemble methods compared with single feature selectors in comparison of stability and accuracy.
  The gray area is the accuracy when using all features.}
  \label{fig:boxplot_svm}
\end{figure}

@@

\appendix
\label{appendix}
\section{Python libraries}
\label{apd:libraries}
For the feature selection methods SU and RELIEFF the scikit-feature library \citep{Li-etal16} is used.
For C-SVM, RFE, Lasso logistic regression the scikit-sklearn library \citep{scikit-learn} is used.
\section{Additional plots}
\label{apd:plots}
\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{images/feature_weights_hist.png}
  \caption{The histogram of the feature weights for all feature selectors used for all datasets used.}
  \label{fig:feature_weights_hist}
\end{figure}
